A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-10-17 model.tex
rough edits for youdin's comments

In this section we review the basic properties of the 
single-fluid mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations. A variable $f$ is
subject to Eulerian perturbations of the form 
\begin{align}
 \real\left[ \delta f(r,z)\exp{\left(-\ii\sigma t\right)},\right]
\end{align}
where $\sigma$ is the complex mode frequency. In Appendix
\ref{var_prin} we derive the integral relation,  
\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.\label{int_rel}
\end{align} 
The real coefficients $A,B,C,D$ can be read off 
Eq. \ref{integral_ex}. Note that $B=C$ from dynamical equilibrium
(Eq. \ref{vshear}). We now consider various limits of
Eq. \ref{int_rel}.

@@

Our thermodynamic interpretation does not rigorously explain
\emph{why} dust drag causes a pressure lag behind density, but merely shows
that this \emph{must} be the case for any growing modes associated the
dust drag. To understand how dust-drag causes this lag requires an
explicit solution to the linearized equations with detailed treatment
of the function $\mathcal{C}$. However, given the complexity of
$\mathcal{C}$ (see Appendix \ref{lin_dust}), we might generally expect
the system to  support a range of stable and unstable modes, with the
latter being associated with pressure-density lag.

>>>>

In this section we review the basic properties of the 
single-fluid mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations. A variable $f$ is
subject to Eulerian perturbations of the form 
\begin{align}
 \real\left[ \delta f(r,z)\exp{\left(-\ii\sigma t\right)}\right], 
\end{align}
where $\sigma$ is the complex mode frequency. In Appendix
\ref{var_prin} we derive the integral relation,  
\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\Bigl\lvert \nabla\cdot\left(P\dd
  \bm{v}\right)\Bigr\rvert^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.\label{int_rel}
\end{align} 
The real coefficients $A,B,C,D$ can be read off 
Eq. \ref{integral_ex}. Note that $B=C$ from dynamical equilibrium
(Eq. \ref{vshear}). We now consider various limits of
Eq. \ref{int_rel}.

@@

Our thermodynamic interpretation does not explain
\emph{why} dust drag causes a pressure lag behind density, but merely shows
that this \emph{must} be the case for any growing modes associated the
dust drag. To rigorously understand how dust-drag causes this lag requires an
explicit solution to the linearized equations with detailed treatment
of the function $\mathcal{C}$. However, given the complexity of
$\mathcal{C}$ (see Appendix \ref{lin_dust}), we might generally expect
the dusty disks to  support a range of stable and unstable modes, with the
latter being associated with pressure-density lag.

### 2016-10-17 result.tex
tiny changes
rough edits for youdin's comments

Fig. \ref{compare_vshear_fixHd} compares the basic state
vertical shear rate, which is destabilizing, and the vertical buoyancy
frequency, which is stabilizing. For the nearly dust-free disk
$\epsilon_0=10^{-3}$ the vertical shear dominates buoyancy for all
$|z|>0$. However, a heavy dust-load with $\epsilon_0=1$ renders the 
buoyancy to dominate over vertical shear in the disk atmosphere 
$|z|\gtrsim 2.5$. We thus expect instability all heights for 
$\epsilon_0=10^{-3}$, but to be restricted to the midplane for
$\epsilon_0=1$.

@@

>>>>

Fig. \ref{compare_vshear_fixHd} compares the basic state
vertical shear rate, which is destabilizing, and the vertical buoyancy
frequency, which is stabilizing. For the nearly dust-free disk
$\epsilon_0=10^{-3}$ the vertical shear dominates buoyancy for all
$|z|>0$. However, a heavy dust-load with $\epsilon_0=1$ renders the 
buoyancy to dominate over vertical shear in the disk atmosphere 
$|z|\gtrsim 2.5\Hg$. We thus expect instability all heights for 
$\epsilon_0=10^{-3}$, but to be restricted to the midplane for
$\epsilon_0=1$.

@@

\subsection{Axisymmetric stability of ultra-thin dust layers}
In \S\ref{iso_perfect} we asserted that in the limit of perfectly
coupled dust (as considered here) that radially smooth, strictly
isothermal disks are stable against axisymmetric perturbations, no
matter how thin the dust layer is. We now demonstrate this
numerically.

For comparison with previous studies of dusty disks, we also use 
the Richardson number $\rich \equiv N_z^2/\left(r\p_z\Omega\right)^2$
to label the calculations here \citep{youdin02}.   
Numerical simulations show   
non-axisymmetric instabilities develops when $\rich\lesssim0.1$, brought about
by very thin dust layers \citep{chiang08, lee10}. We show that \emph{axisymmetric}
instabilities never develop, however small $\rich$.

We shall consider ultra thin dust layers with $\Hd\leq 0.01\Hg$ and thus
restrict the vertical domain to $\zmax = 0.02\Hg$. We fix the 
metalicity $Z=0.01$ so the midpane dust-to-gas ratio $\epsilon_0$ is 
$O(1)$.  We consider radial wavenumbers such $k\Hd=1$.

Fig. \ref{ultra_thin} shows the maximum growth rate  as a function of
the radial temperature gradient, $q$. For all cases the vertical
shear is dominated by that due to the dust layer (see
\S\ref{vertshear}). However, we see that growth rates $s\propto
|q|$. In particular, this holds for $\rich < 0.1$, the critical value
for non-axisymmetric instabilities.

Axisymmetric instability is associated with the thermal
contribution to vertical shear: as $q\to0$,  $\p_z\Omega$ becomes
entirely due to dust-settling and there is no instability ($s\to
0$). This result is independent of $\rich$, so the Richardson number
is not useful in characterizing the axisymmetric stability of dust
layers.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_thindust} 
  \caption{Maximum VSI growth rate for ultra-thin dust layers $\Hd
    \leq 0.01\Hg$. Vertical shear is dominated by that due to
    dust-settling, but axisymmetric instability is associated with the
    radial temperature gradient $q$.
    \label{ultra_thin}
    }
\end{figure}
